---
title: "PPSI v4: Diagnostics, Sensitivity and Hotspot Analysis"
subtitle: "Western Ireland | 500 m spatial resolution"
author:
  - "Justin Lingg-Laham"
  - "Jonah van den Bos"
lang: en
format:
  html:
    toc: true
    embed-resources: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

## Analysis overview

This document contains the post-processing, diagnostic and robustness analysis
for the **Pulse-Pollution Susceptibility Index (PPSI)** developed for western
Ireland. The spatial model combines three normalized components:

\[
\mathrm{PPSI\text{-}P} = \sqrt[3]{S \cdot M \cdot C}
\]

where \(S\) is agricultural source presence, \(M\) is mobilization potential
and \(C\) is hydrological connectivity. The geometric mean produces high index
values only where all three components provide support.

The analysis has five objectives:

1. verify the distribution and internal consistency of the 500-m PPSI raster;
2. translate the continuous index into relative pixel-priority classes;
3. calculate mean PPSI scores and rankings for 76 HydroBASINS catchments;
4. test whether alternative aggregation rules and component weights alter the
   catchment priorities;
5. identify catchments that repeatedly occur among the five highest-ranked
   catchments.

Three spatial products must be distinguished:

- **Priority pixels** are 500-m cells classified from the regional PPSI
  distribution. The highest class represents the upper 1% of agriculturally
  relevant pixels.
- **Candidate output zones** are interpreted spatial clusters of priority
  pixels. They are useful for communication and follow-up investigation but are
  not calculated as part of the sensitivity analysis.
- **Robust hotspot catchments** repeatedly occur among the five highest mean
  catchment scores across the tested model variants. A *core hotspot* occurs in
  the Top 5 under all five variants.

The sensitivity analysis is an **internal robustness check**. It tests whether
the spatial priorities depend strongly on reasonable modelling choices. It is
not an external validation against measured pollutant concentrations or
observed pollution events.

## 1. Setup

```{r}
#| label: setup

required_packages <- c(
  "terra",
  "ggplot2",
  "dplyr",
  "tidyr",
  "patchwork",
  "scales",
  "sf",
  "knitr"
)

missing_packages <- required_packages[
  !vapply(
    required_packages,
    requireNamespace,
    logical(1),
    quietly = TRUE
  )
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(terra)
library(ggplot2)
library(dplyr)
library(tidyr)
library(patchwork)
library(scales)
library(sf)

options(
  scipen = 999,
  digits = 4
)

theme_set(
  theme_minimal(base_size = 12)
)

project_colour <- "#002060"
accent_colour <- "#C51B7D"
secondary_text <- "#617086"
```

### Input and output paths

The default path follows the repository structure. Set the environment variable
`PPSI_DATA_DIR` if the exported files are stored elsewhere.

```{r}
#| label: paths

data_folder <- Sys.getenv(
  "PPSI_DATA_DIR",
  unset = file.path(
    "data",
    "derived_data",
    "v4"
  )
)

raster_file <- file.path(
  data_folder,
  "WestIreland_PPSI_P_v4_story_500m_merged.tif"
)

csv_file <- file.path(
  data_folder,
  "WestIreland_PPSI_P_v4_statistics_500m.csv"
)

catchment_vector_file <- file.path(
  data_folder,
  "WestIreland_PPSI_P_v4_catchments_500m.geojson"
)

presentation_folder <- file.path(
  data_folder,
  "presentation_outputs"
)

dir.create(
  presentation_folder,
  recursive = TRUE,
  showWarnings = FALSE
)

if (!file.exists(raster_file)) {
  stop("Raster file not found: ", raster_file)
}

if (!file.exists(csv_file)) {
  stop("Catchment statistics file not found: ", csv_file)
}

if (file.info(csv_file)$mtime < file.info(raster_file)$mtime) {
  warning(
    paste(
      "The catchment CSV is older than the raster export.",
      "Confirm that both files represent the same model version."
    )
  )
}
```

## 2. Raster import

The current Google Earth Engine export contains 22 bands. The first seven bands
retain the names and order of the former story export, so all downstream object
and layer names remain unchanged. The import also accepts the legacy seven-band
export for backwards compatibility.

```{r}
#| label: raster-import

story_band_names <- c(
  "source_presence",
  "rainfall_pulse",
  "soil_runoff",
  "mobilization",
  "connectivity",
  "ppsi_continuous",
  "ppsi_quintile"
)

history_band_names <- c(
  "cropland_fraction",
  "cultivated_grass_fraction",
  "mean_annual_rx1day_mm",
  "heavy_rain_days_per_year",
  "rainfall_intensity_0_1",
  "heavy_rain_frequency_0_1",
  "soil_texture_usda_class",
  "elevation_mean_m",
  "slope_deg",
  "hand_m",
  "stream_proxy_500m",
  "distance_to_stream_m",
  "slope_factor_0_1",
  "stream_proximity_0_1",
  "low_hand_factor_0_1"
)

all_band_names <- c(
  story_band_names,
  history_band_names
)

combined_raster <- rast(raster_file)

if (nlyr(combined_raster) == length(all_band_names)) {

  names(combined_raster) <- all_band_names

} else if (nlyr(combined_raster) == length(story_band_names)) {

  names(combined_raster) <- story_band_names

  warning(
    paste(
      "A legacy seven-band story export was loaded.",
      "The core analysis remains available, but the additional model-history",
      "bands are not present."
    )
  )

} else {

  stop(
    "The raster contains ",
    nlyr(combined_raster),
    " bands. Expected 7 or 22 bands."
  )
}

# Preserve the object and names used by the original analysis.
ppsi_raster <- combined_raster[[story_band_names]]

print(ppsi_raster)

cat(
  "\nRaster resolution:",
  paste(res(ppsi_raster), collapse = " × "),
  "m\n",
  "Number of exported bands:",
  nlyr(combined_raster),
  "\n"
)
```

## 3. Agriculturally relevant pixels

Only complete land pixels with at least 5% agricultural source presence are
included in the PPSI diagnostics. This reproduces the source gate applied in
the Google Earth Engine model.

```{r}
#| label: pixel-data

minimum_source_fraction <- 0.05

analysis_band_names <- c(
  "source_presence",
  "rainfall_pulse",
  "soil_runoff",
  "mobilization",
  "connectivity",
  "ppsi_continuous"
)

pixel_data_all <- as.data.frame(
  ppsi_raster[[analysis_band_names]],
  na.rm = FALSE
)

pixel_data_all <- pixel_data_all[
  complete.cases(pixel_data_all),
]

pixel_data <- pixel_data_all[
  pixel_data_all$source_presence >= minimum_source_fraction &
    pixel_data_all$ppsi_continuous >= 0 &
    pixel_data_all$ppsi_continuous <= 1,
]

if (nrow(pixel_data) == 0) {
  stop("No agriculturally relevant PPSI pixels were found.")
}

ppsi_values <- pixel_data$ppsi_continuous

component_columns <- c(
  "source_presence",
  "mobilization",
  "connectivity"
)

component_names <- c(
  source_presence = "Agricultural source",
  mobilization = "Mobilization",
  connectivity = "Connectivity"
)

cat(
  "Complete land pixels:", nrow(pixel_data_all), "\n",
  "Agriculturally relevant pixels:", nrow(pixel_data), "\n",
  "Share retained:",
  round(100 * nrow(pixel_data) / nrow(pixel_data_all), 1),
  "%\n"
)
```

## 4. Raster diagnostics

### Continuous PPSI distribution

```{r}
#| label: ppsi-summary

summary_probabilities <- c(
  0,
  0.05,
  0.10,
  0.25,
  0.50,
  0.75,
  0.80,
  0.90,
  0.95,
  0.99,
  1
)

distribution_table <- data.frame(
  Statistic = c(
    "Minimum",
    "5th percentile",
    "10th percentile",
    "25th percentile",
    "Median",
    "75th percentile",
    "80th percentile",
    "90th percentile",
    "95th percentile",
    "99th percentile",
    "Maximum",
    "Mean",
    "Standard deviation"
  ),
  PPSI = c(
    as.numeric(
      quantile(
        ppsi_values,
        probs = summary_probabilities,
        na.rm = TRUE
      )
    ),
    mean(ppsi_values, na.rm = TRUE),
    sd(ppsi_values, na.rm = TRUE)
  )
)

distribution_table$PPSI <- round(
  distribution_table$PPSI,
  3
)

knitr::kable(
  distribution_table,
  caption = "Distribution of the continuous PPSI v4."
)
```

```{r}
#| label: fig-ppsi-distribution
#| fig-cap: "Distribution of the continuous PPSI across agriculturally relevant 500-m cells."
#| fig-width: 8
#| fig-height: 4.5

ppsi_distribution_plot <- ggplot(
  data.frame(PPSI = ppsi_values),
  aes(x = PPSI)
) +
  geom_histogram(
    binwidth = 0.025,
    boundary = 0,
    fill = "#2C7FB8",
    color = "white",
    linewidth = 0.2
  ) +
  geom_vline(
    xintercept = median(ppsi_values),
    color = "#222222",
    linewidth = 0.9,
    linetype = "dashed"
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1)
  ) +
  labs(
    x = "Continuous PPSI",
    y = "Number of 500-m cells"
  )

ppsi_distribution_plot
```

### Component distributions

```{r}
#| label: component-summary

component_summary <- do.call(
  rbind,
  lapply(
    component_columns,
    function(component) {

      component_values <- pixel_data[[component]]

      data.frame(
        Component = unname(component_names[component]),
        Mean = mean(component_values, na.rm = TRUE),
        Median = median(component_values, na.rm = TRUE),
        P10 = quantile(component_values, 0.10, na.rm = TRUE),
        P90 = quantile(component_values, 0.90, na.rm = TRUE)
      )
    }
  )
)

component_summary[c("Mean", "Median", "P10", "P90")] <- round(
  component_summary[c("Mean", "Median", "P10", "P90")],
  3
)

knitr::kable(
  component_summary,
  caption = "Distribution of the three normalized PPSI components."
)
```

```{r}
#| label: fig-component-distributions
#| fig-cap: "Distributions of source presence, mobilization and hydrological connectivity."
#| fig-width: 9
#| fig-height: 4.5

component_long <- stack(
  pixel_data[component_columns]
)

names(component_long) <- c(
  "Value",
  "Component"
)

component_long$Component <- factor(
  component_long$Component,
  levels = component_columns,
  labels = unname(component_names)
)

component_distribution_plot <- ggplot(
  component_long,
  aes(
    x = Value,
    fill = Component
  )
) +
  geom_histogram(
    binwidth = 0.025,
    boundary = 0,
    color = "white",
    linewidth = 0.15
  ) +
  facet_wrap(
    ~ Component,
    nrow = 1
  ) +
  scale_x_continuous(
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2)
  ) +
  scale_fill_manual(
    values = c(
      "#D95F02",
      "#7570B3",
      "#1B9E77"
    ),
    guide = "none"
  ) +
  labs(
    x = "Normalized component value",
    y = "Number of cells"
  )

component_distribution_plot
```

### Component support for high PPSI values

The following diagnostic tests whether cells in the upper regional PPSI fifth
receive support from all three components rather than from only one dominant
component.

```{r}
#| label: high-value-support

upper_fifth_threshold <- as.numeric(
  quantile(
    ppsi_values,
    0.80,
    na.rm = TRUE
  )
)

high_pixels <- pixel_data$ppsi_continuous >= upper_fifth_threshold

component_matrix <- as.matrix(
  pixel_data[component_columns]
)

minimum_component <- apply(
  component_matrix,
  1,
  min,
  na.rm = TRUE
)

high_support_table <- data.frame(
  Statistic = c(
    "Threshold of the upper PPSI fifth",
    "Median source value",
    "Median mobilization value",
    "Median connectivity value",
    "Median of the weakest component",
    "High-PPSI cells with any component below 0.30 (%)"
  ),
  Value = c(
    upper_fifth_threshold,
    median(pixel_data$source_presence[high_pixels], na.rm = TRUE),
    median(pixel_data$mobilization[high_pixels], na.rm = TRUE),
    median(pixel_data$connectivity[high_pixels], na.rm = TRUE),
    median(minimum_component[high_pixels], na.rm = TRUE),
    100 * mean(minimum_component[high_pixels] < 0.30, na.rm = TRUE)
  )
)

high_support_table$Value <- round(
  high_support_table$Value,
  3
)

knitr::kable(
  high_support_table,
  caption = "Component support within the upper regional PPSI fifth."
)
```

## 5. Relative pixel-priority classes

The continuous PPSI is converted into relative regional priority classes using
thresholds calculated only from agriculturally relevant pixels. These classes
support visualization and spatial screening; they are not absolute pollution
risk classes.

```{r}
#| label: priority-raster

priority_breaks <- as.numeric(
  quantile(
    ppsi_values,
    probs = c(
      0.80,
      0.90,
      0.95,
      0.99
    ),
    na.rm = TRUE
  )
)

names(priority_breaks) <- c(
  "p80",
  "p90",
  "p95",
  "p99"
)

continuous_raster <- ppsi_raster[["ppsi_continuous"]]
source_raster <- ppsi_raster[["source_presence"]]

# Class 0: less than 5% agricultural source presence
# Class 1: lower 80% of relevant pixels
# Class 2: upper 20%
# Class 3: upper 10%
# Class 4: upper 5%
# Class 5: upper 1%
priority_raster <- ifel(
  is.na(continuous_raster) | is.na(source_raster),
  NA,
  ifel(
    source_raster < minimum_source_fraction,
    0,
    ifel(
      continuous_raster < priority_breaks["p80"],
      1,
      ifel(
        continuous_raster < priority_breaks["p90"],
        2,
        ifel(
          continuous_raster < priority_breaks["p95"],
          3,
          ifel(
            continuous_raster < priority_breaks["p99"],
            4,
            5
          )
        )
      )
    )
  )
)

names(priority_raster) <- "PPSI_relative_priority"

priority_output <- file.path(
  data_folder,
  "WestIreland_PPSI_P_v4_priority_percentiles_500m_new.tif"
)

writeRaster(
  priority_raster,
  filename = priority_output,
  overwrite = TRUE,
  datatype = "INT1U",
  NAflag = 255,
  gdal = c(
    "COMPRESS=DEFLATE",
    "TILED=YES"
  )
)

priority_frequency <- freq(priority_raster)

print(priority_breaks)
print(priority_frequency)

cat("Priority raster saved to:\n", priority_output, "\n")
```

## 6. Catchment sensitivity analysis

For each model variant, the mean PPSI value of every catchment is ranked from
highest to lowest. The base-model ranking is compared with four alternatives:

- an additive arithmetic model;
- a source-weighted geometric model;
- a mobilization-weighted geometric model;
- a connectivity-weighted geometric model.

Spearman's rank correlation evaluates the complete ordering of all catchments.
Top-5 overlap evaluates whether the highest-priority catchments remain selected.

### Catchment data and model definitions

```{r}
#| label: catchment-import

catchments <- read.csv(
  csv_file,
  stringsAsFactors = FALSE,
  check.names = FALSE,
  colClasses = "character"
)

catchments$HYBAS_ID <- as.character(
  catchments$HYBAS_ID
)

scenario_columns <- c(
  Base = "PPSI_P_base_geometric",
  Additiv = "PPSI_P_additive_sensitivity",
  Quelle = "PPSI_P_source_heavy",
  Mobilisierung = "PPSI_P_mobilization_heavy",
  Konnektivitaet = "PPSI_P_connectivity_heavy"
)

missing_columns <- setdiff(
  unname(scenario_columns),
  names(catchments)
)

if (length(missing_columns) > 0) {
  stop(
    "Missing columns in the catchment CSV: ",
    paste(missing_columns, collapse = ", ")
  )
}

catchments[unname(scenario_columns)] <- lapply(
  catchments[unname(scenario_columns)],
  as.numeric
)

if ("SUB_AREA" %in% names(catchments)) {
  catchments$SUB_AREA <- as.numeric(catchments$SUB_AREA)
}

model_variants <- data.frame(
  Modellvariante = names(scenario_columns),
  Model = c(
    "Base model (our PPSI)",
    "Additive model",
    "Source weighted",
    "Mobilization weighted",
    "Connectivity weighted"
  ),
  Source = c(33, 33, 50, 25, 25),
  Mobilization = c(33, 33, 25, 50, 25),
  Connectivity = c(33, 33, 25, 25, 50),
  Combination = c(
    "Geometric mean",
    "Arithmetic mean",
    "Weighted geometric mean",
    "Weighted geometric mean",
    "Weighted geometric mean"
  ),
  Formula = c(
    "(S × M × C)^(1/3)",
    "(S + M + C) / 3",
    "S^0.50 × M^0.25 × C^0.25",
    "S^0.25 × M^0.50 × C^0.25",
    "S^0.25 × M^0.25 × C^0.50"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  model_variants[
    c(
      "Model",
      "Combination",
      "Formula"
    )
  ],
  caption = "Model variants used in the sensitivity analysis."
)

cat("Catchments imported:", nrow(catchments), "\n")
```

### Rank stability and Top-5 overlap

```{r}
#| label: sensitivity-results

rank_columns <- paste0(
  "rank_",
  names(scenario_columns)
)

for (i in seq_along(scenario_columns)) {

  score_column <- unname(scenario_columns[i])
  rank_column <- rank_columns[i]

  catchments[[rank_column]] <- rank(
    -catchments[[score_column]],
    ties.method = "min",
    na.last = "keep"
  )
}

top5_list <- lapply(
  scenario_columns,
  function(score_column) {

    valid_order <- order(
      catchments[[score_column]],
      decreasing = TRUE,
      na.last = NA
    )

    catchments$HYBAS_ID[
      head(valid_order, 5)
    ]
  }
)

base_top5 <- top5_list[["Base"]]

comparison_rows <- lapply(
  names(scenario_columns)[-1],
  function(variant_name) {

    score_column <- unname(
      scenario_columns[variant_name]
    )

    rank_test <- suppressWarnings(
      cor.test(
        catchments[[scenario_columns["Base"]]],
        catchments[[score_column]],
        method = "spearman",
        exact = FALSE
      )
    )

    data.frame(
      Modellvariante = variant_name,
      Gemeinsame_Top5 = length(
        intersect(
          base_top5,
          top5_list[[variant_name]]
        )
      ),
      Spearman_Rho = unname(rank_test$estimate),
      P_Wert = rank_test$p.value
    )
  }
)

comparison_base <- do.call(
  rbind,
  comparison_rows
)

comparison_display <- comparison_base
comparison_display$Spearman_Rho <- round(
  comparison_display$Spearman_Rho,
  3
)
comparison_display$P_Wert <- format.pval(
  comparison_display$P_Wert,
  digits = 3,
  eps = 0.001
)

knitr::kable(
  comparison_display,
  col.names = c(
    "Model variant",
    "Top-5 overlap",
    "Spearman rho",
    "p-value"
  ),
  caption = paste(
    "Sensitivity of the catchment ranking relative to the",
    "equal-weighted geometric base model."
  )
)
```

The p-value evaluates the null hypothesis of no monotonic association between
the two catchment rankings. It does not measure effect size and does not provide
external validation. Rank stability is primarily interpreted from the very high
Spearman coefficients together with the Top-5 overlap.

## 7. Robust hotspot catchments

The robustness score counts how often each catchment appears among the five
highest-ranked catchments across the base model and four alternatives.

```{r}
#| label: robust-hotspots

rank_matrix <- as.matrix(
  catchments[rank_columns]
)

catchments$Mittlerer_Rang <- rowMeans(
  rank_matrix,
  na.rm = TRUE
)

catchments$Rang_SD <- apply(
  rank_matrix,
  1,
  sd,
  na.rm = TRUE
)

catchments$Bester_Rang <- apply(
  rank_matrix,
  1,
  min,
  na.rm = TRUE
)

catchments$Schlechtester_Rang <- apply(
  rank_matrix,
  1,
  max,
  na.rm = TRUE
)

catchments$Top5_Haeufigkeit <- rowSums(
  rank_matrix <= 5,
  na.rm = TRUE
)

robustness_order <- order(
  -catchments$Top5_Haeufigkeit,
  catchments$Mittlerer_Rang,
  catchments$Rang_SD
)

output_columns <- c(
  "HYBAS_ID",
  "Mittlerer_Rang",
  "Rang_SD",
  "Bester_Rang",
  "Schlechtester_Rang",
  "Top5_Haeufigkeit"
)

if ("SUB_AREA" %in% names(catchments)) {
  output_columns <- append(
    output_columns,
    "SUB_AREA",
    after = 1
  )
}

robust_hotspots <- catchments[
  robustness_order,
  output_columns
]

robust_hotspots$Mittlerer_Rang <- round(
  robust_hotspots$Mittlerer_Rang,
  1
)

robust_hotspots$Rang_SD <- round(
  robust_hotspots$Rang_SD,
  1
)

robust_hotspots$Robustness_Class <- ifelse(
  robust_hotspots$Top5_Haeufigkeit == 5,
  "Core hotspot: 5/5",
  ifelse(
    robust_hotspots$Top5_Haeufigkeit == 4,
    "Stable hotspot: 4/5",
    ifelse(
      robust_hotspots$Top5_Haeufigkeit >= 1,
      "Scenario-sensitive candidate",
      "Not selected"
    )
  )
)

hotspot_candidates <- robust_hotspots[
  robust_hotspots$Top5_Haeufigkeit >= 1,
]

hotspot_candidates <- hotspot_candidates[
  order(
    -hotspot_candidates$Top5_Haeufigkeit,
    hotspot_candidates$Mittlerer_Rang,
    hotspot_candidates$Rang_SD
  ),
]

hotspot_candidates$Map_Label <- paste0(
  "H",
  seq_len(nrow(hotspot_candidates))
)

number_catchments <- nrow(robust_hotspots)
number_in_any_top5 <- sum(robust_hotspots$Top5_Haeufigkeit >= 1)
number_in_four_top5 <- sum(robust_hotspots$Top5_Haeufigkeit >= 4)
number_in_all_top5 <- sum(robust_hotspots$Top5_Haeufigkeit == 5)

knitr::kable(
  head(robust_hotspots, 10),
  caption = "The ten most robust catchment priorities."
)

cat(
  "Catchments in at least one Top 5:", number_in_any_top5, "\n",
  "Catchments in at least four Top 5 lists:", number_in_four_top5, "\n",
  "Core hotspots present in all five Top 5 lists:", number_in_all_top5, "\n"
)
```

## 8. Presentation figures

Only two figures are retained for presentation and repository use: a compact
sensitivity figure and a hotspot-frequency figure. All values are calculated
from the imported catchment results rather than entered manually.

```{r}
#| label: presentation-theme

presentation_theme <- theme_minimal(
  base_size = 14,
  base_family = "sans"
) +
  theme(
    plot.title = element_text(
      color = project_colour,
      face = "bold",
      size = 18
    ),
    plot.subtitle = element_text(
      color = secondary_text,
      size = 12
    ),
    plot.caption = element_text(
      color = secondary_text,
      size = 9,
      hjust = 0
    ),
    axis.title = element_text(
      color = project_colour,
      face = "bold"
    ),
    axis.text = element_text(
      color = "#26364A"
    ),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom",
    plot.margin = margin(12, 18, 12, 12)
  )
```

### Compact sensitivity figure

```{r}
#| label: fig-sensitivity-compact
#| fig-cap: "Tested model variants and their effects on catchment rankings."
#| fig-width: 14
#| fig-height: 6.3

model_order <- rev(model_variants$Model)

model_variants$Model <- factor(
  model_variants$Model,
  levels = model_order
)

weight_data <- model_variants |>
  pivot_longer(
    cols = c(
      Source,
      Mobilization,
      Connectivity
    ),
    names_to = "Component",
    values_to = "Weight"
  ) |>
  mutate(
    Component_x = match(
      Component,
      c(
        "Source",
        "Mobilization",
        "Connectivity"
      )
    )
  )

weights_plot <- ggplot(
  weight_data,
  aes(
    x = Component_x,
    y = Model
  )
) +
  geom_tile(
    aes(fill = Weight),
    width = 0.86,
    height = 0.68,
    color = "white",
    linewidth = 1.2
  ) +
  geom_text(
    aes(
      label = paste0(round(Weight), "%"),
      color = Weight >= 45
    ),
    size = 4.7,
    fontface = "bold"
  ) +
  geom_text(
    data = model_variants,
    aes(
      x = 4.05,
      y = Model,
      label = Combination
    ),
    inherit.aes = FALSE,
    hjust = 0,
    color = secondary_text,
    size = 3.4
  ) +
  scale_fill_gradientn(
    colors = c(
      "#E8EEF7",
      "#B8B2D2",
      accent_colour
    ),
    values = scales::rescale(c(25, 33, 50)),
    limits = c(25, 50)
  ) +
  scale_color_manual(
    values = c(
      "FALSE" = project_colour,
      "TRUE" = "white"
    )
  ) +
  scale_x_continuous(
    breaks = 1:3,
    labels = c(
      "Source presence",
      "Mobilization",
      "Connectivity"
    ),
    limits = c(0.5, 5.25)
  ) +
  labs(
    title = "[A] Model variants",
    subtitle = "Component weights and aggregation method",
    x = NULL,
    y = NULL
  ) +
  presentation_theme +
  theme(
    axis.text.x = element_text(
      color = project_colour,
      face = "bold",
      size = 10
    ),
    axis.text.y = element_text(
      color = project_colour,
      face = "bold",
      size = 10
    ),
    panel.grid = element_blank(),
    legend.position = "none"
  )

ranking_plot_data <- data.frame(
  Modellvariante = c(
    "Base",
    comparison_base$Modellvariante
  ),
  Spearman_Rho = c(
    1,
    comparison_base$Spearman_Rho
  ),
  Gemeinsame_Top5 = c(
    NA,
    comparison_base$Gemeinsame_Top5
  ),
  P_Wert = c(
    NA,
    comparison_base$P_Wert
  )
)

ranking_plot_data <- merge(
  ranking_plot_data,
  model_variants[c("Modellvariante", "Model")],
  by = "Modellvariante",
  all.x = TRUE,
  sort = FALSE
)

ranking_plot_data$Model <- factor(
  ranking_plot_data$Model,
  levels = model_order
)

ranking_plot_data$Result_label <- ifelse(
  ranking_plot_data$Modellvariante == "Base",
  "Reference",
  sprintf(
    "ρ = %.3f     Top 5 = %d/5",
    ranking_plot_data$Spearman_Rho,
    ranking_plot_data$Gemeinsame_Top5
  )
)

ranking_plot <- ggplot(
  ranking_plot_data,
  aes(
    y = Model,
    x = Spearman_Rho
  )
) +
  geom_vline(
    xintercept = 1,
    color = project_colour,
    linewidth = 0.8,
    linetype = "dashed"
  ) +
  geom_segment(
    aes(
      x = 0.978,
      xend = Spearman_Rho,
      yend = Model
    ),
    color = "#CED6E1",
    linewidth = 2
  ) +
  geom_point(
    aes(
      color = Modellvariante == "Base"
    ),
    size = 4.8
  ) +
  geom_text(
    aes(
      x = 1.002,
      label = Result_label
    ),
    hjust = 0,
    color = "#24344A",
    size = 3.7
  ) +
  scale_color_manual(
    values = c(
      "FALSE" = accent_colour,
      "TRUE" = "#8E9AA9"
    )
  ) +
  scale_x_continuous(
    breaks = c(0.98, 0.99, 1.00),
    limits = c(0.977, 1.025)
  ) +
  labs(
    title = "[B] Effect on catchment priorities",
    subtitle = "Comparison with the base-model ranking",
    x = "Spearman rank correlation",
    y = NULL
  ) +
  presentation_theme +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.x = element_line(color = "#E1E6ED"),
    legend.position = "none"
  )

all_significant <- all(
  comparison_base$P_Wert < 0.001,
  na.rm = TRUE
)

sensitivity_figure <- weights_plot +
  ranking_plot +
  plot_layout(widths = c(1.25, 1)) +
  plot_annotation(
    title = "Catchment priorities remain stable across model variants",
    subtitle = paste0(
      "Mean PPSI values for ",
      nrow(catchments),
      " catchments; ",
      ifelse(
        all_significant,
        "all alternative rank correlations p < 0.001.",
        "p-values are reported in the sensitivity table."
      )
    ),
    caption = paste(
      "High rank correlation and Top-5 overlap indicate internal robustness,",
      "not validation against observed pollution data."
    ),
    theme = theme(
      plot.title = element_text(
        color = project_colour,
        face = "bold",
        size = 20
      ),
      plot.subtitle = element_text(
        color = secondary_text,
        size = 12
      ),
      plot.caption = element_text(
        color = secondary_text,
        size = 9,
        hjust = 0
      ),
      plot.background = element_rect(
        fill = "white",
        color = NA
      )
    )
  )

print(sensitivity_figure)

ggsave(
  filename = file.path(
    presentation_folder,
    "06_sensitivity_analysis_compact.png"
  ),
  plot = sensitivity_figure,
  width = 14,
  height = 6.3,
  units = "in",
  dpi = 400,
  bg = "white"
)
```

### Hotspot robustness figure

```{r}
#| label: fig-hotspot-robustness
#| fig-cap: "Frequency with which candidate catchments occur among the five highest-ranked catchments."
#| fig-width: 10
#| fig-height: 5.5

hotspot_plot_data <- hotspot_candidates

hotspot_plot_data$Map_Label <- factor(
  hotspot_plot_data$Map_Label,
  levels = rev(hotspot_plot_data$Map_Label)
)

robustness_colors <- c(
  "Core hotspot: 5/5" = accent_colour,
  "Stable hotspot: 4/5" = "#F28E2B",
  "Scenario-sensitive candidate" = "#D9B44A"
)

hotspot_frequency_plot <- ggplot(
  hotspot_plot_data,
  aes(
    x = Top5_Haeufigkeit,
    y = Map_Label,
    fill = Robustness_Class
  )
) +
  geom_col(width = 0.62) +
  geom_text(
    aes(
      label = paste0(Top5_Haeufigkeit, "/5")
    ),
    hjust = -0.18,
    color = "#26364A",
    fontface = "bold",
    size = 4.5
  ) +
  scale_fill_manual(values = robustness_colors) +
  scale_x_continuous(
    limits = c(0, 5.75),
    breaks = 0:5,
    expand = c(0, 0)
  ) +
  labs(
    title = paste0(
      number_in_all_top5,
      " catchments remain in the Top 5 in every model"
    ),
    subtitle = paste0(
      number_in_any_top5,
      " of ",
      number_catchments,
      " catchments appear in at least one Top-5 list."
    ),
    x = "Number of model variants with a Top-5 rank",
    y = NULL,
    fill = NULL,
    caption = paste(
      "Core hotspot is a project-specific robustness definition,",
      "not an externally validated pollution class."
    )
  ) +
  presentation_theme +
  theme(
    axis.text.y = element_text(
      face = "bold",
      color = project_colour
    )
  )

hotspot_frequency_plot

ggsave(
  filename = file.path(
    presentation_folder,
    "03_hotspot_robustness.png"
  ),
  plot = hotspot_frequency_plot,
  width = 10,
  height = 5.5,
  units = "in",
  dpi = 320,
  bg = "white"
)
```

## 9. Optional spatial hotspot export

If the HydroBASINS catchment polygons are available, the robustness results are
joined by `HYBAS_ID` and exported as a GeoPackage for cartographic styling in
QGIS. No interactive file selection is used, which keeps the workflow
reproducible.

```{r}
#| label: hotspot-spatial-export

if (file.exists(catchment_vector_file)) {

  catchment_polygons <- st_read(
    catchment_vector_file,
    quiet = TRUE
  )

  catchment_polygons$HYBAS_ID <- as.character(
    catchment_polygons$HYBAS_ID
  )

  catchments_robust <- catchment_polygons |>
    left_join(
      robust_hotspots,
      by = "HYBAS_ID"
    ) |>
    mutate(
      Top5_Haeufigkeit = coalesce(
        Top5_Haeufigkeit,
        0
      ),
      Robustness_Class = coalesce(
        Robustness_Class,
        "Not selected"
      )
    )

  hotspot_output <- file.path(
    presentation_folder,
    "WestIreland_PPSI_P_v4_hotspot_robustness.gpkg"
  )

  st_write(
    catchments_robust,
    hotspot_output,
    layer = "catchment_robustness",
    delete_dsn = TRUE,
    quiet = TRUE
  )

  cat("QGIS hotspot layer saved to:\n", hotspot_output, "\n")

} else {

  message(
    paste(
      "Catchment polygons were not found.",
      "The tabular hotspot analysis was completed, but the optional",
      "GeoPackage was not created."
    )
  )
}
```

## 10. Export summary tables

```{r}
#| label: export-tables

write.csv(
  distribution_table,
  file.path(
    data_folder,
    "PPSI_v4_distribution_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  component_summary,
  file.path(
    data_folder,
    "PPSI_v4_component_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  high_support_table,
  file.path(
    data_folder,
    "PPSI_v4_high_value_support.csv"
  ),
  row.names = FALSE
)

write.csv(
  comparison_base,
  file.path(
    data_folder,
    "PPSI_v4_sensitivity_summary.csv"
  ),
  row.names = FALSE
)

write.csv(
  robust_hotspots,
  file.path(
    data_folder,
    "PPSI_v4_robust_hotspots.csv"
  ),
  row.names = FALSE
)

presentation_key_numbers <- data.frame(
  Metric = c(
    "Catchments analysed",
    "Catchments in at least one Top-5 list",
    "Catchments in at least four Top-5 lists",
    "Catchments in all five Top-5 lists",
    "Minimum Spearman rank correlation",
    "Maximum Spearman rank correlation"
  ),
  Value = c(
    number_catchments,
    number_in_any_top5,
    number_in_four_top5,
    number_in_all_top5,
    min(comparison_base$Spearman_Rho, na.rm = TRUE),
    max(comparison_base$Spearman_Rho, na.rm = TRUE)
  )
)

write.csv(
  presentation_key_numbers,
  file.path(
    presentation_folder,
    "PPSI_v4_presentation_key_numbers.csv"
  ),
  row.names = FALSE
)

knitr::kable(
  presentation_key_numbers,
  caption = "Key results exported for the presentation and repository."
)
```

## Interpretation

High Spearman rank correlations indicate that the overall ordering of the
catchments changes little across the alternative formulations. Top-5 overlap
adds a decision-oriented check by showing whether the catchments selected as
the highest priorities remain the same. Catchments that occur in the Top 5
under all five formulations are therefore robust to the tested assumptions.

This stability supports the use of the PPSI as a **relative spatial screening
tool**. It does not prove that the index predicts measured pollution correctly.
That claim would require independent water-quality or event-observation data.

